---
title: "Practical Machine Learning Assignment"
author: "Alana Souter"
date: "28 November 2017"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Overview

The goal of this project is to predict the manner in which the 6 participants did the exercise. This is the "classe" variable in the training set. This report describes how the model was built, how cross validation was used and what the expected out of sample error is. 

## The Data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source:
http://groupware.les.inf.puc-rio.br/har

Special thanks go to the authors for allowing the data to be used for this assignment.

**Preparing the Environment**

```{r chunk1}
library(knitr)
library(rpart)
library(rpart.plot)
library(randomForest)
library(caret)
set.seed(12345)
opts_chunk$set(echo=TRUE,results='hold')
```

**Loading the Data**
```{r chunk2}
setwd("D:/Users/soutal/datasciencecoursera")
training <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!",""))
testing <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!",""))
```

**Initial Exploratory Analysis & Cleanup**
```{r chunk3}
dim(training)
dim(testing)
```

```{r chunk4}
str(training,list.len=20)
str(testing,list.len=20)
```


a) Remove variables with too many NA values
```{r chunk6}
allNA<-sapply(training,function(x) mean(is.na(x)))
training<-training[,allNA==FALSE]
dim(training)
allNA<-sapply(testing,function(x) mean(is.na(x)))
testing<-testing[,allNA==FALSE]
dim(testing)
```

b) Check for numeric variables which have no or zero variance to see if we can reduce the variables this way
```{r chunk7}
nearZeroVar(training[, sapply(training,is.numeric)])
nearZeroVar(testing[, sapply(testing,is.numeric)])
```

c) Remove idenfication only variables
```{r chunk8}
training<-training[,-(1:5)]
testing<-testing[,-(1:5)]
```

d) Split data for cross validation into a training and testing set. A 60/40 split has been chosen for training/testing.
```{r chunk9}
inTrain<-createDataPartition(y=training$classe, p=0.60, list=FALSE)
train1<-training[inTrain,]
test1<-training[-inTrain,]
dim(train1)
dim(test1)
```

##Models for Prediction##

**Random Forest**
```{r chunk10}
model1<-randomForest(classe~.,data=train1,ntree=1000)
varImpPlot(model1)
```
We will now check this models accuracy
```{r chunk11}
predictions<-predict(model1,newdata=test1)
confusionMat<-confusionMatrix(predictions,test1$classe)
confusionMat
```

This model has a 99.6% accuracy on the test dataset!  On this basis I will not test any other models (such as Decision Tree) as I'm happy this model is a good option for applying to the predictions.

**Out of sample error rate**
The above Random Forest out of sample error rate is calculated by:
```{r chunk12}
missClass = function(values, predicted) {
  sum(predicted != values) / length(values)
}
OOS_errRate = missClass(test1$classe, predictions)
OOS_errRate
```
The out of sample error rate is 0.34%

##Prediction##

We will now use the Random Forest model on the validation(testing) data.  
From attempting to run this using "prediction<-predict(model1,newdata=testing)" I discovered there was an inconsistency between the training & testing data which created an error message when running randomForest.  In order to find the issue and correct it I carried out the following:
```{r chunk13}
sapply(train1, class)
sapply(testing, class)
```
This highlighted that there were class mismatches  in the columns "magnet_dumbbell_z, magnet_forearm_y and magnet_forearm_z, problem_id".  I therefore had to correct this before running the prediction again:

```{r chunk14}
fixFrame<-head(training,1)
fixFrame<-fixFrame[,-length(colnames(fixFrame))]
validation<-testing[,-(55)]
validation<-rbind(fixFrame,validation)
validation<-validation[-1,]
prediction<-predict(model1,newdata=validation)
prediction
```